---
title: "My_Tidy_Tuesday_2020_Week_24"
author: "Joseph Edgerton"
date: "6/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#----------------------------Load Data--------------------------

```{r loading packages and downloading data}
#Download packages

my_packages <- c("tidyverse", "ggrepel", "here", "devtools", "egg")

install.packages(my_packages, repos = "http://cran.rstudio.com")

#load packages

library(tidyverse)
library(devtools)

#positon_jitterdodge in p1
library(ggrepel)

#ggarange
library(egg)

#Download Data and start to define objects

devtools::install_github("thebioengineer/tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load(2020, week = 24)

firsts <- tuesdata$firsts

science <- tuesdata$science

#set theme
theme_set(theme_bw())

```

#-------------------------view data---------------------------

```{r}
head(firsts)
glimpse(firsts)
firsts

unique(firsts$gender)

view(firsts)

head(science)
glimpse(science)
science

```

OK, a lot of text information. I am not familiar with doing data vis with text so I'll see what I can do.

#--------------data cleaning/ selection------------------

I will work with the firsts data set.

First, I want to code gender into "male" and "female" to clarify the data.
Second, I will remove any NA's from the set since I only want to work with complete data at the moment.

```{r}
Gender_firsts <- firsts %>% 
  mutate(gender = recode(gender, "African-American Firsts" = "male", "Female African American Firsts" = "female")) %>% 
  na.omit()
```

#-----------------plotting---------------------------

I will select and work with the firsts dataset using year, gender and category.

I need to make a graph to visualise the data by category and year.

```{r}
#basic plot for forming ideas
p0 <- ggplot(data = Gender_firsts, aes(x = year, y = category))

p0 + geom_point(aes(color = category)) + guides(color = FALSE)


```

Now, Let's do two things:
1. color by male/Female
2. jitter the points to space them out

```{r}
#need to assign colors to each gender

Gender_firsts$gender <- factor(Gender_firsts$gender, levels = c("male", "female"))
gender_colors <- c("deepskyblue", "violetred2")

#plot

p1 <- ggplot(data = Gender_firsts, aes(x = year, y = category))

p1 + geom_point(aes(color = gender), position = position_jitterdodge()) +
  guides(color = FALSE) + 
  scale_color_manual(values = gender_colors)

```

Next, I want to try to add a few horizontal lines across each category to visually separate the categories.

```{r}
p1 <- ggplot(data = Gender_firsts, aes(x = year, y = category))

p1.1 <- p1 + geom_point(aes(color = gender), position = position_jitterdodge()) +
  guides(color = FALSE) + 
  scale_color_manual(values = gender_colors) +
  geom_hline(yintercept = 1.5:7.5, size = 1.25) + 
  geom_hline(yintercept = 1:8) +labs(x = NULL)

p1.1
```

Now, we can place a column geom on the y-axis to show the percent of male and female in each category.

The plot may be getting a little busy at the moment, but let's see how this goes.


After viewing p0, and later observing the breakdown in category by gender over time (p1), I want to create a plot that examines the percent female and male observations for a category.

```{r}
#getting the percent of male and female per category
Gender_firsts_per <- Gender_firsts %>%   
  group_by(category, gender) %>% 
  summarize(n = n()) %>% 
  mutate(gender_freq = n/sum(n),
         gender_per = round((gender_freq*100), 0))

#assign colors to the new data set for columns

Gender_firsts_per$gender <- factor(Gender_firsts_per$gender, levels = c("male", "female"))
gender_colors <- c("deepskyblue", "violetred2")

#generate new column geom, add % text and adjust scales to %

p2 <- ggplot(data = Gender_firsts_per, aes(x = category, y = gender_freq,
                                              fill = gender))
p2.1 <- p2  + geom_col(position = "dodge") + 
  scale_fill_manual(values = gender_colors) +
  coord_flip() + 
  geom_text(mapping = aes(label = paste0(gender_per, "%"), hjust = 1),
                           position = position_dodge(width = 1)) +
  scale_y_continuous(labels = scales::percent) +
  geom_vline(xintercept = 1.5:7.5, size = 1.25) + geom_vline(xintercept = 1:8) +
  labs(y = NULL)

p2.1
```

I will now use the egg package to combine the p1.1 (geom_point) and p2.1 (geom_col) plots.

```{r final plot}
#egg package's ggarrange for final plot

p3 <- ggarrange(p2.1 +
               theme(axis.ticks.y = element_blank(),
                     plot.margin = margin(r = 1),
                     panel.grid.major.x = element_blank(),
                     panel.grid.minor.x = element_blank()) +
            labs(x = "Category of Accomplishment", fill = "Gender") +
              labs(title = "African-Amercian First Achievements Sorted by Gender",
                   subtitle = "Achievement category composition and time plots (1738-2019)") +
            theme(legend.position = c(-0.5, 0.97)) +
            guides(fill = guide_legend(reverse=TRUE)), 
          p1.1 +
            labs(caption = "Data source: firsts.csv, Tidy Tuesday Github Week 24, 2020") +
               theme(axis.text.y = element_blank(),
                     axis.ticks.y = element_blank(),
                     axis.title.y = element_blank(),
                     plot.margin = margin(r = 3),
                     panel.grid.major.x = element_blank(),
                     panel.grid.minor.x = element_blank()),
          nrow = 1, widths = c(1,2)) 

```

```{r output plot}

#produce final plot

ggsave(filename = "My_Tidy_Tuesday_2020_Week_24.png", plot = p3, width = 7, height = 5,
       units = "in")
```


Result:

Now, I can examine M/F diversity in each of the categories.


Thoughts:

As my first data vis project this was a cool experience. I got to solve problems and work through design issues along the way.

I decided to only do the "firsts" data set. Within the set, I focused on gender for no particular reason.

I am excited to do more TidyTuesdays in the future. I hope to learn a lot about R and data visualizations along the way.

Thank you


