---
title: "My_Tidy_Tuesday_2020_week_28"
author: "Joseph Edgerton"
date: "7/6/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#----------------------------set up--------------

```{r loading packages and downloading data}
#download packages
my_packages <- c("tidyverse", "tidytuesdayR")

install.packages(my_packages, repos = "http://cran.rstudio.com")

#load packages

library(tidyverse)
library(tidytuesdayR)

#get data from github

tuesdata <- tidytuesdayR::tt_load(2020, week = 28)

#rename data

coffee_table <- tuesdata$coffee_ratings

#set the theme
theme_set(theme_light())

```

###---------------------view data------------------------------------

```{r exploring and viewing data}
#look at data

glimpse(coffee_table)

hist(coffee_table$total_cup_points, breaks = 50)

summary(coffee_table$total_cup_points)

```


#--------------------------tidying-------------------

IDEA: I will focus on comparing the bean processing method to the cup points. Also, I intend to use point size to signify case counts for additional comparisons.

To start, I will: 1. create smaller table, 2. Form groups by processing_method, 3.Find the median cup points per process method, 4. Add the Category one and two defects into a new coluumn and find the median for each method (not sure if going to include).

```{r data process table}
clean_coffee_table <- coffee_table %>% 
  group_by(processing_method) %>% 
  select(total_cup_points, processing_method, category_one_defects, category_two_defects) %>% 
  mutate(median_total_c_points = median(total_cup_points)) %>% 
  mutate(median_total_defects_per_method = median(category_two_defects + category_one_defects)) %>% 
  mutate(count = n()) %>% 
  na.omit()


clean_coffee_table
```


#-------------------------plotting-------------------

```{r plot}
#get x axis displaying methods by greatest score value in descending order
p1 <- ggplot(data = clean_coffee_table, aes(x = reorder(processing_method, median_total_c_points), y = median_total_c_points))


#need to make 3 geom_points for coffee cups
p1.5 <- p1 + geom_point(pch = 22, col = "black", bg = "white", size = 3,
                position = position_nudge(x = 0, y = -0.0625)) +
  geom_point(pch = 21, size = 12, col = "black", bg = "white") +
  geom_point(aes(size = count), color = "#3c2a1e") + 
  
#flipping axes, and manually defining legend and score axis
  coord_flip() + scale_size_continuous("# of judged\nsubmissions\n(range):",range = c(4,10), breaks = c(50, 350, 800)) +
  ylim(81.5, 83) +
  
#styling plot with color and moving labels/text around the plot
  theme(panel.background = element_rect(fill = "tan"),
        plot.background = element_rect(fill = "#f1e8dc"),
        legend.position = c(0.2,0.65),
        legend.box.background = element_rect(color = "black", size = 1.1),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(color = "#f1e8dc"),
        axis.text = element_text(color = "#3c2a1e"),
        axis.title = element_text(face = "bold"),
        text = element_text(color = "#3c2a1e"),
        plot.title = element_text(hjust = 1),
        plot.subtitle = element_text(hjust = 1),
        plot.caption = element_text(hjust = -2.8)) +
  xlab(label = "Coffee bean processing method\n") +
  ylab(label = "\nMedian coffee score (0-100)\n") +
  labs(title = "Average brewed coffee scores for each coffee bean processing method",
       subtitle = "Fairly small range of coffee scores between bean process methods",
          caption = "Data: coffee_ratings.csv, Tidy Tuesday, Week 28, 2020 | By: Joseph Edgerton")

p1.5

```


```{r saving plot}
#save plot
ggsave(path = "figures", filename = "My_Tidy_Tuesday_week_28_2020.png", plot = p1.5,
       width = 7, height = 5)

```






###Extra, not used

```{r combined defect count table}

#was going to do something with defects, but decided not to
defect_table_check <- clean_coffee_table %>% 
  group_by(processing_method,median_total_defects_per_method) %>%
  summarize()
defect_table_check

```





