---
title: "My_Tidy_Tuesday_Week_25_2020"
author: "Joseph Edgerton"
date: "6/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#----------------------------set up--------------

```{r loading packages and downloading data}
#download packages
my_packages <- c("tidyverse", "devtools")

install.packages(my_packages, repos = "http://cran.rstudio.com")

#load packages

library(tidyverse)
library(devtools)

#get data from github
devtools::install_github("thebioengineer/tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load(2020, week = 25)

#create the individual data set

census <- tuesdata$census

#set the theme
theme_set(theme_light())
```

#--------------------------------view data------------------------------------

Right now, I think I will work with the census data.

```{r}
#look at data

glimpse(census)

```

#--------------------------tidying/plotting (a little sporadic)-------------------

I have an idea for a stacked bar plot for each division of the US over time. 


A new table with race and values separated, each having percentages.

```{r}
#now I will pivot the data to race and race_count columns that both contain white and black populations within a specific area and year

#then I will calculate the relative frequency and percent for the white and black observations

#also na.omit since I want to work with complete data

pct_wb_census <- census %>% 
  pivot_longer(c(-region, -division, -year, -black_free, -black_slaves, -total),
               names_to = "race", values_to = "race_count" ) %>% 
  group_by(region, division, year) %>% 
  mutate(freq = race_count/sum(race_count),
         pct = round((freq*100),0)) %>% 
  na.omit()


pct_wb_census

```

Need to see how a stacked bar plot over time looks

```{r}
#test plot of stacking with respect to time

p0 <- ggplot(data = pct_wb_census, aes(x = year, y = pct, fill = race))

p0 + geom_col(position = "fill")

```

#############
Quick test to investigate totals for columns in creating frequencies and percents
```{r table test}
#test to see if "Total column is total for black and white populations and if "black column is total for free and slaves.

test_census <- census %>% 
  mutate(test_of_total = (white + black) - total) %>% 
  mutate(test_of_black_pop = (black_free + black_slaves) - black)

test_census

#the total column does not represent ONLY the black and white popualtions in this data set

#the black column does represent a total for both black_free and black_slaves

#I will use this infomration going forward
```

################
Not bad, now I need to create a new table that also pivots the black_free and black_slaves columns into a column "status" and "status_count".

pct_fs_census = "Percent of ___ free or enslaved black population ___ from census data" 

The current idea is to have a stacked bar plot showing relative pct of white and black population (not using the "total" column since the earlier test showed extra, uncategorized observations) using the race_count column

Then, within the black population section of the bar have a color correspond to the percent of the black populaiton that was free within the area and time.

```{r}
#new table pivoting black_free and black_slaves into one column with values in another column. 

#Also, creating different thresholds for colors that represent the percent of the free black population (using case_when)

#only using a small subset with a filter

pct_fs_census <- pct_wb_census %>%
  pivot_longer(c(-region, -division, -year, -race, -race_count,
                 -freq, -pct, -total),
               names_to = "status", values_to = "status_count" ) %>% 
  group_by(region, division, year, race) %>%
  mutate(freq_fs = status_count/sum(status_count),
         pct_fs = round((freq_fs*100),0)) %>% 
  filter(region == "Northeast", division == "New England") %>% 
  mutate(pct_free = case_when(status == "black_free"~ pct_fs)) %>% 
  na.omit() %>% 
  mutate(fill_col = case_when(
  race == "white" ~ "NA",
  race == "black" & status == "black_free" & pct_free == 0 ~ "blue",
  race == "black" & status == "black_free" & ((pct_free < 25) & (pct_free >= 1)) ~ "green",
  race == "black" & status == "black_free" & ((pct_free < 50) & (pct_free >= 25))  ~ "yellow",
  race == "black" & status == "black_free" & ((pct_free < 75) & (pct_free >= 50)) ~ "orange",
  race == "black" & status == "black_free" & ((pct_free < 100) & (pct_free >= 75)) ~ "red",
  race == "black" & status == "black_free" & pct_free == 100 ~ "purple"
  ))

pct_fs_census

#the variable pct_free may be redundant, but it was a way to focus on the black_free observations and not display the black_slaves observations (yet still count them in the relative percent of the free black and black slave populations)

#I need to create factors from the color character column and tell R to order the colors on top of the non-colored (NA) bar that will represent the white population

pct_fs_census$fill_col <- factor(pct_fs_census$fill_col)

pct_fs_census$fill_col <- fct_rev(pct_fs_census$fill_col)

```


Now time to plot the New England division and see the how the test colors show on the plot 

```{r}
#test plot with colors representing percent of the free black population in the total black population

#scaling the x-axis as a character to display all the years.

#coloring the bar outline by race (which I don't know if it does anything besides helping to organize bars by race? Does fill already do that? IDK)

#using fill to color the top section with scale_fill_identity filling with color and NA (transparent)

p1 <- ggplot(data = pct_fs_census,
             aes(x = year, y = pct,
                 color = race_status,
                 fill = fill_col)) 

p1 + geom_col(position = "fill", color = "black") +
  scale_fill_identity(guide = "legend") + 
  scale_y_continuous(labels = scales::percent, expand = c(0,0)) +
  theme(plot.margin = margin(r = 3),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(title = "filler") +
  scale_x_continuous("year",labels = as.character(pct_fs_census$year), breaks = pct_fs_census$year)


```

The plotting looks good. Now I need to create a new table for all divisons (facet_wrap) with slightly better(scaled) colors.

```{r}
#using 6 of the viridis colors to make a color scale (may not be the best choice, but I wanted to try it out)

final_pct_fs_census <- pct_wb_census %>%
  pivot_longer(c(-region, -division, -year, -race, -race_count,
                 -freq, -pct, -total),
               names_to = "status", values_to = "status_count" ) %>% 
  group_by(region, division, year, race) %>%
  mutate(freq_fs = status_count/sum(status_count),
         pct_fs = round((freq_fs*100),0)) %>% 
  mutate(pct_free = case_when(status == "black_free"~ pct_fs)) %>% 
  na.omit() %>% 
  mutate(fill_col = case_when(
  race == "white" ~ "NA",
  race == "black" & status == "black_free" & pct_free == 0 ~ "#440154FF",
  race == "black" & status == "black_free" & ((pct_free < 25) & (pct_free >= 1)) ~ "#404788FF",
  race == "black" & status == "black_free" & ((pct_free < 50) & (pct_free >= 25))  ~ "#238A8DFF",
  race == "black" & status == "black_free" & ((pct_free < 75) & (pct_free >= 50)) ~ "#55C667FF",
  race == "black" & status == "black_free" & ((pct_free < 100) & (pct_free >= 75)) ~ "#B8DE29FF",
  race == "black" & status == "black_free" & pct_free == 100 ~ "#FDE725FF"
  ))

final_pct_fs_census

```


```{r}
#test colors on plot with all divisions (facet_wrap)
p2 <- ggplot(data = final_pct_fs_census,
             aes(x = year, y = pct,
                 color = race_status,
                 fill = fill_col)) 

p2 + geom_col(position = "fill", color = "black") +
  scale_fill_identity(guide = "legend") + 
  scale_y_continuous(labels = scales::percent) +
  theme(plot.margin = margin(r = 3),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(title = "filler") +
  scale_x_continuous("year",labels = as.character(pct_fs_census$year), breaks = pct_fs_census$year) + facet_wrap(~division)

```

OK, way too small to distinguish details within certain divisions. I should make the plot a little bigger.

Final plot:
```{r final, fig.width=12, fig.height=10}
# I want to order columns in facet_wrap by west, midwest  and northeast and then have the bottom row be the south (kind of the order you would see on a map)
final_pct_fs_census$division <- factor(final_pct_fs_census$division,
                                       levels = c("Pacific", "West North Central",
                                                  "New England", "Mountain",
                                                  "East North Central",
                                                  "Middle Atlantic",
                                                  "West South Central",
                                                  "East South Central",
                                                  "South Atlantic"))

#Also, the legend needs to have an order to it. This factor will place the color represetning "100%" at the top of the ordering.

final_pct_fs_census$fill_col <- factor(final_pct_fs_census$fill_col,
                                       levels = c("#FDE725FF","#B8DE29FF",
                                                  "#55C667FF","#238A8DFF",
                                                  "#404788FF","#440154FF", "NA"))
#plot

p3 <- ggplot(data = final_pct_fs_census,
             aes(x = year, y = pct,
                 color = race_status,
                 fill = fill_col )) 

#I have added a descriptive title and some labels to my legend. Also I added a distinct box around the legend

#I made the facet strip background and axis text black for contrast.

#Lastly, I got rid of the y-axis (the unit of percent is already apparent) and re-labeled the x-axis. I also created a descriptive title and caption.

p4 <- p3 + geom_col(position = "fill", color = "black") +
  scale_fill_identity("Race\n(Black population\nhighlighted by\n% free)",
                      labels = c("100% free", "75-99% free",
                                "50-74% free", "25-49% free",
                                "1-24% free", "White population"),
                      guide = "legend") + 
  scale_y_continuous(labels = scales::percent) +
  theme(plot.margin = margin(r = 3),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background =element_rect(fill="black"),
        axis.text.y = element_text(color = "black"),
        axis.text.x = element_text(color = "black"),
        legend.box.background = element_rect(color="black", size=1),
        legend.position = c(0.08,0.86)) +
  labs(title = "Percentage of Total Black and White Populations by Census Year with Breakdown of Free Black Population (1790-1870)", caption = "Data from: census.csv, Tidy Tuesday, Week 25, 2020 | By: Joseph Edgerton", x = "Year", y = NULL) +
  scale_x_continuous(labels = as.character(pct_fs_census$year), breaks = pct_fs_census$year) + facet_wrap(~division)

p4
```

Notes:

I wanted to put extra information in a stacked bar chart without cluttering the plot. I think I gave that idea a good try. I'm not sure if the information is easy to take-away from the plot.

Either way, I learned about a lot of new functions and arguments (pivot, scale_fill_identity, case_when, etc.).

Still, at this scale it is hard to see all the data. I could make individual plots but I would like to be able to compare them (e.g. by division).

Also, the color scale was based on some arbitraty colors from the viridis color scale. I'm still trying to figure out colors and scales and how to properly use them.

Thank you.

```{r}
#save image

ggsave(filename = "My_Tidy_Tuesday_week_25_2020.png", plot = p4,
       width = 12, height = 10)

```




